\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-05-22 jss/main.tex
combine section 2 and 3; rearrange section 3 with bullet points

The rest of the paper is organized as follows: Section \ref{cube} introduces the proposed cube structure as a way to conceptualize multivariate spatio-temporal data. Section \ref{cubble} presents the main design and functionality of the \pkg{cubble} package. Section \ref{others} explains how the \pkg{cubble} package deals with more advanced considerations, including data with hierarchical structure, data matching and how the package fits with existing static and interactive visualization tools. Moreover we also illustrate how the \pkg{cubble} package deals with spatio-temporal data transformations. Section \ref{examples} uses Australian weather station data and river level data as examples to demonstrate the use of the package. An example of how the \pkg{cubble} package handles Network Common Data Form (NetCDF) data is also provided. Section \ref{conclude} discuss the paper contributions and future directions.

\hypertarget{cube}{%
\section{Conceptual framework: spatio-temporal cube}\label{cube}}

Spatio-temporal data can be conceptualized using a cubical data model with three axes which typically are, time, latitude and longitude. This abstraction can be useful for generalizing operations and visualization purposes: \citet{lu_multidimensional_2018} shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; \citet{bach_review_2014} reviews the temporal data visualization based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data'', but refers to ``an abstract 2D substrate that is used to visualize data at a specific time''. Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualizations, still applies to our spatio-temporal data manipulation and visualization approaches.

The most common space-time cube uses the three axes, time, latitude, longitude, and can be considered stacking space across time. Ours is a multivariate spatio-temporal cube with the three axes defined to be time, site and variables, as illustrated in the leftmost column of Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hyper-cubes when describing multivariate spatio-temporal data and is the conceptual framework behind the \code{cubble} class. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.

While the data cube model is conceptually convenient for spatio-temporal data, a 3D data array is not sufficiently rich for data wrangling, for several reasons. Although arrays can be efficient for the computation on numerical values, spatio-temporal data typically includes various types of variables. For example, character strings and specific datetime classes are common. In addition, it will be generally useful to be able to create new variables which is trickier to manage in an array. Thus for convenient wrangling, we have opted to create a special \code{cubble} class.

{\centering \includegraphics[width=1\linewidth]{../figures/diagram-keynotes/diagram-keynotes.001}

\caption[An illustration of the conceptual spatio-temporal cube with different slicing on time, site, and variable]{An illustration of the conceptual spatio-temporal cube with different slicing on time, site, and variable. For each axis, the slicing can be on a single value or a set of values.}\label{fig:cube-def}
\end{figure}
\end{CodeChunk}

@@

@@

\hypertarget{create}{%
\subsection{Create a cubble}\label{create}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_nested <- climate_flat |>
+   as_cubble(key = id, index = date, coords = c(long, lat))
R> cubble_nested
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>  
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_nested |>
+   mutate(rain_day = sum(ts$prcp != 0))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name          wmo_id ts       rain_day
  <chr>       <dbl> <dbl> <dbl> <chr>          <dbl> <list>      <int>
1 ASN00009021 -31.9  116.  15.4 perth airport  94610 <tibble>      104
2 ASN00010311 -31.9  117. 179   york           94623 <tibble>       89
3 ASN00010614 -32.9  117. 338   narrogin       94627 <tibble>       90
4 ASN00014015 -12.4  131.  30.4 darwin airpo~  94120 <tibble>      106
5 ASN00015131 -17.6  134. 220   elliott        94236 <tibble>       63
\end{CodeOutput}
\end{CodeChunk}

@@

\hypertarget{change-focus-by-facing-the-time-variables}{%
\subsection{Change focus by facing the time-variables}\label{change-focus-by-facing-the-time-variables}}

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). The \code{cubble} class also provides a long form, which expands the \code{ts} column and temporarily ``hides'' the spatial variables. The function \code{face_temporal()} is used to switch from the nested \code{cubble} into the long one. The first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the \code{cubble} object just created into its long form:

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_long <- cubble_nested |> face_temporal()
R> cubble_long
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id
#   [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

\hypertarget{change-focus-back-to-the-site-variable-face}{%
\subsection{Change focus back to the site-variable face}\label{change-focus-back-to-the-site-variable-face}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_back <- cubble_long |> face_spatial()
R> cubble_back
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>  
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble>
\end{CodeOutput}
\begin{CodeInput}
R> identical(cubble_nested, cubble_back)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{unfold}{%
\subsection{Unfold spatial variables into the long form}\label{unfold}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_unfold <- cubble_long |> unfold(long, lat)
R> cubble_unfold
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id
#   [dbl]
  id          date        prcp  tmax  tmin  long   lat
  <chr>       <date>     <dbl> <dbl> <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3  116. -31.9
2 ASN00009021 2020-01-02     0  24.9  16.4  116. -31.9
3 ASN00009021 2020-01-03     6  23.2  13    116. -31.9
4 ASN00009021 2020-01-04     0  28.4  12.4  116. -31.9
5 ASN00009021 2020-01-05     0  35.3  11.6  116. -31.9
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{tidyverse}{%
\subsection{Why not just use the existing tidyverse functions}\label{tidyverse}}

\hypertarget{compact}{%
\subsection{Compatibility with existing packages}\label{compact}}

The \pkg{cubble} package leverages tools available in existing packages used for spatial and temporal analysis, specifically, \pkg{dplyr}, \pkg{tsibble}, \pkg{sf} (\pkg{s2}), and \pkg{ncdf4}, as explained here.

\subsubsection[]{\code{dplyr}}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> climate_flat_ts <- climate_flat |> 
+   tsibble::as_tsibble(key = id, index = date)
R> climate_flat_cb <-  climate_flat_ts |> 
+   cubble::as_cubble(coords = c(long, lat))
R> climate_flat_cb
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>  
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> climate_flat_cb |> 
+   mutate(fabletools::features(
+     ts, tmax, list(tmax_mean = mean, tmax_var = var)
+     ))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id          lat  long  elev name  wmo_id ts       tmax_mean tmax_var
  <chr>     <dbl> <dbl> <dbl> <chr>  <dbl> <list>       <dbl>    <dbl>
1 ASN00009~ -31.9  116.  15.4 pert~  94610 <tbl_ts>      25.7    38.6 
2 ASN00010~ -31.9  117. 179   york   94623 <tbl_ts>      26.2    51.1 
3 ASN00010~ -32.9  117. 338   narr~  94627 <tbl_ts>      23.7    45.4 
4 ASN00014~ -12.4  131.  30.4 darw~  94120 <tbl_ts>      33.1     3.02
5 ASN00015~ -17.6  134. 220   elli~  94236 <tbl_ts>      34.6    24.7 
\end{CodeOutput}
\end{CodeChunk}

@@

@@

\subsubsection[]{\code{ncdf4}}\label{netcdf}

@@

Linking is also available starting from the time series plot, by selecting points. This will activate rows having the same id in the long \code{cubble}. The corresponding rows in the nested \code{cubble} are activated, and highlighted in the map. (An illustration can be found in the appendix.) Note that this type of linking, both from the map or the time series, is what \citet{CS07} would call categorical variable linking, where station id is the categorical variable.

@@

\hypertarget{examples}{%
\section{Examples}\label{examples}}

@@

Glyph maps work well to explore temporal patterns across spatial locations, particularly when the spatial locations are gridded. In this example, they are irregularly spaced, which can result in overlapping glyphs obscuring each other. To fix this, one could aggregate data from nearby stations. An example of this is included in the Appendix.

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.52, -37.73, 146.06, -36.55]
# temporal: date [date], matched_var [dbl]
  id          name        lat  long type   dist group ts       n_match
  <chr>       <chr>     <dbl> <dbl> <chr> <dbl> <int> <list>     <int>
1 405234      SEVEN CR~ -36.9  146. river  6.15     5 <tibble>      21
2 404207      HOLLAND ~ -36.6  146. river  8.54    10 <tibble>      21
3 ASN00082042 strathbo~ -36.8  146. clim~  6.15     5 <tibble>      21
4 ASN00082170 benalla ~ -36.6  146. clim~  8.54    10 <tibble>      21
5 230200      MARIBYRN~ -37.7  145. river  6.17     6 <tibble>      19
# ... with 3 more rows
\end{CodeOutput}
\end{CodeChunk}

>>>>

The rest of the paper is organized as follows: Section \ref{cubble} presents the main design and functionality of the \pkg{cubble} package. Section \ref{others} explains how the \pkg{cubble} package deals with more advanced considerations, including data with hierarchical structure, data matching and how the package fits with existing static and interactive visualization tools. Moreover we also illustrate how the \pkg{cubble} package deals with spatio-temporal data transformations. Section \ref{examples} uses Australian weather station data and river level data as examples to demonstrate the use of the package. An example of how the \pkg{cubble} package handles Network Common Data Form (NetCDF) data is also provided. Section \ref{conclude} discuss the paper contributions and future directions.

@@

\hypertarget{the-cubble-object}{%
\subsection{The cubble object}\label{the-cubble-object}}

\emph{the conceptual framework + nested and long form as per the design vignette}

\begin{itemize}
\tightlist
\item
  diagram,
\item
  two forms: long/ nested - spatial/ temporal
\item
  attributes
\item
  headers
\end{itemize}

@@

\hypertarget{create}{%
\subsection{Casting into the cubble class}\label{create}}

\begin{itemize}
\tightlist
\item
  key, index, coords
\item
  import different data: an expand data, spatial + temporal, netcdf, stars, sftime
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_nested <- climate_flat |>
+   as_cubble(key = id, index = date, coords = c(long, lat))
R> cubble_nested
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.9764, -32.9342, 133.5407, -12.4239]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>  
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_nested |>
+   mutate(rain_day = sum(ts$prcp != 0))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.9764, -32.9342, 133.5407, -12.4239]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts       rain_~1
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>     <int>
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble>     104
2 ASN00010311 -31.9  117. 179   york            94623 <tibble>      89
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble>      90
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble>     106
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble>      63
# ... with abbreviated variable name 1: rain_day
\end{CodeOutput}
\end{CodeChunk}

@@

\emph{netcdf}

\hypertarget{functions-and-methods-in-cubble}{%
\subsection{Functions and methods in cubble}\label{functions-and-methods-in-cubble}}

\emph{Print all the methods with methods(class = ``cubble\_df'')}

\hypertarget{cubble-specific-methods}{%
\subsubsection{cubble specific methods}\label{cubble-specific-methods}}

The three cubble specific functions are \code{face_temporal()}, \code{face_spatial()}, and \code{unfold}.

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). The \code{cubble} class also provides a long form, which expands the \code{ts} column and stores the spatial variables as an attribute. The function \code{face_temporal()} is used to switch from the nested \code{cubble} into the long one. The first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the \code{cubble} object just created into its long form:

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_long <- cubble_nested |> face_temporal()
R> cubble_long
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.9764, -32.9342, 133.5407, -12.4239]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id
#   [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_back <- cubble_long |> face_spatial()
R> cubble_back
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.9764, -32.9342, 133.5407, -12.4239]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>  
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble>
\end{CodeOutput}
\begin{CodeInput}
R> identical(cubble_nested, cubble_back)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_unfold <- cubble_long |> unfold(long, lat)
R> cubble_unfold
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.9764, -32.9342, 133.5407, -12.4239]
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id
#   [dbl]
  id          date        prcp  tmax  tmin  long   lat
  <chr>       <date>     <dbl> <dbl> <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3  116. -31.9
2 ASN00009021 2020-01-02     0  24.9  16.4  116. -31.9
3 ASN00009021 2020-01-03     6  23.2  13    116. -31.9
4 ASN00009021 2020-01-04     0  28.4  12.4  116. -31.9
5 ASN00009021 2020-01-05     0  35.3  11.6  116. -31.9
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{dplyr-methods}{%
\subsubsection{dplyr methods}\label{dplyr-methods}}

@@

\hypertarget{compatibility-with-other-packages}{%
\subsection{Compatibility with other packages}\label{compatibility-with-other-packages}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> climate_flat_ts <- climate_flat |> 
+   tsibble::as_tsibble(key = id, index = date)
R> climate_flat_cb <-  climate_flat_ts |> 
+   cubble::as_cubble(coords = c(long, lat))
R> climate_flat_cb
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.9764, -32.9342, 133.5407, -12.4239]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>  
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tbl_ts>
2 ASN00010311 -31.9  117. 179   york            94623 <tbl_ts>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tbl_ts>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tbl_ts>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tbl_ts>
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> climate_flat_cb |> 
+   mutate(fabletools::features(
+     ts, tmax, list(tmax_mean = mean, tmax_var = var)
+     ))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.9764, -32.9342, 133.5407, -12.4239]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name   wmo_id ts       tmax_~1 tmax_~2
  <chr>       <dbl> <dbl> <dbl> <chr>   <dbl> <list>     <dbl>   <dbl>
1 ASN00009021 -31.9  116.  15.4 perth~  94610 <tbl_ts>    25.7   38.6 
2 ASN00010311 -31.9  117. 179   york    94623 <tbl_ts>    26.2   51.1 
3 ASN00010614 -32.9  117. 338   narro~  94627 <tbl_ts>    23.7   45.4 
4 ASN00014015 -12.4  131.  30.4 darwi~  94120 <tbl_ts>    33.1    3.02
5 ASN00015131 -17.6  134. 220   ellio~  94236 <tbl_ts>    34.6   24.7 
# ... with abbreviated variable names 1: tmax_mean, 2: tmax_var
\end{CodeOutput}
\end{CodeChunk}

@@

\emph{add an example on changing CRS?}

@@

\hypertarget{tidyverse}{%
\subsection{Comparison to other spatio-temporal classes}\label{tidyverse}}

From the initial \pkg{spacetime} \citep{spacetime} package. New packages, such as \pkg{stars} \citep{stars} and \pkg{sftime} \citep{sftime}, have {[}\ldots{]}. This section compares and contrast the cubble data structure with other existing alternative, namely \pkg{stars} and \pkg{sftime}.

The \texttt{stars} package creates a \texttt{stars} object from an array and a dimension object. In many cases, i.e.~satellite images, this is convenient, however, there are also many spatio-temporal data are formatted in 2D data frames.

create a dimension object, populate the data into the array

\begin{itemize}
\tightlist
\item
  array in \proglang{R} allows for character,
\item
  specific datetime class
\end{itemize}

\emph{make the example simpler}

\begin{itemize}
\tightlist
\item
  \texttt{temporal\$var1\ \%\textgreater{}\%\ matrix(byrow\ =\ TRUE,\ nrow\ =\ 2,\ ncol\ =\ 5)}
\end{itemize}

This happens when different stations have different start date and require other software to reshape the data correctly.

The approach taken by the \pkg{sftime} package is to combine the spatial and temporal variables together into a joint table. {[}for sf objects with daily observation, memory efficiency{]}

Consider the climate data used in the \pkg{cubble} package, \code{climate_aus}. It contains daily precipitation, minimum and maximum temperature for 639 weather stations across Australia in 2020. The converted sftime object is 13.6830016 times the size of the cubble object:

\begin{CodeChunk}
\begin{CodeInput}
R> climate_aus %>% sftime::st_as_sftime() %>% object.size()
\end{CodeInput}
\begin{CodeOutput}
118243136 bytes
\end{CodeOutput}
\begin{CodeInput}
R> climate_aus %>% object.size()
\end{CodeInput}
\begin{CodeOutput}
8641608 bytes
\end{CodeOutput}
\end{CodeChunk}

@@

Linking is also available starting from the time series plot, by selecting points. This will activate rows having the same id in the long \code{cubble}. The corresponding rows in the nested \code{cubble} are activated, and highlighted on the map. (An illustration can be found in the appendix.) Note that this type of linking, both from the map or the time series, is what \citet{CS07} would call categorical variable linking, where station id is the categorical variable.

@@

\hypertarget{examples}{%
\section{Applications}\label{examples}}

@@

Glyph maps work well to explore temporal patterns across spatial locations, particularly when the spatial locations are gridded. In this example, they are irregularly spaced, which can result in overlapping glyphs obscuring each other. To fix this, one could aggregate data from nearby stations. An example of this use is included in the Appendix.

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.5203, -37.727706, 146.059205, -36.5523]
# temporal: date [date], matched_var [dbl]
  id          name        lat  long type   dist group ts       n_match
  <chr>       <chr>     <dbl> <dbl> <chr> <dbl> <int> <list>     <int>
1 230200      MARIBYRN~ -37.7  145. river  6.17     6 <tibble>      19
2 404207      HOLLAND ~ -36.6  146. river  8.54    10 <tibble>      21
3 405234      SEVEN CR~ -36.9  146. river  6.15     5 <tibble>      21
4 406213      CAMPASPE~ -37.0  145. river  1.84     1 <tibble>      18
5 ASN00082042 strathbo~ -36.8  146. clim~  6.15     5 <tibble>      21
# ... with 3 more rows
\end{CodeOutput}
\end{CodeChunk}
